Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-05 paper/paper.tex
derivation of KL in comment
compressed the section on bayesian methods, located missing citations, fixed the plot in appendix

Deep neural networks are an integral part of the machine learning and data science set of
tools for practical data-driver problem solving. Despite recent advances in generalization
and optimization theory specific to deep networks, it is challenging to deploy the trained
in actual embedded hardware, due to obvious storage and compute limitations. Many techniques
exist to compress or sparsify the networks: weight pruning \cite{citation_needed}, quantization
from IEEE754 floating point to integer-based arithmetic, and dropout. The key idea is that
the more explicitly zero parameters the model has, the less multiplications the mode needs.

Bayesian Inference is a principled framework of reasoning about uncertainty. The core
principle is updating prior beliefs in accordance with the likelihood of given empirical
observations into posterior belief. Under suitable modelling and prior assumptions, methods
of variational Bayesian inference can be used towards model sparsification: variants of
Bayesian dropout studied in \cite{kingma_variational_2015,molchanov_variational_2017}, and
\cite{kharitonov_variational_2018}.

@@

In general, the core of Bayesian Inference can be summarized as follows: given a prior
distribution $\pi(m)$ on models (hypotheses) $\mathcal{M}$, utilize the empirical evidence
$D$ to update the assumptions by considering the likelihood of the observations under each
$m \in \mathcal{M}$. The inference is performed using the Bayes rule to get the exact
posterior belief: $
  p(m \mid D) = \tfrac{p(D \mid m) \pi(m)}{p(D)}
$ with $
  p(D) = \mathbb{E}_{\pi(m)} p(D \mid m)
$.

It is usually assumed that $\mathcal{M}$ is a parametric family indexed by a parameter $\omega$
from some set $\Omega$, that each model is \textit{identified} with its parameter, and that
each $m = m_\omega(\cdot)$ specifies the functional form of the conditional distribution of
the data $D = (z_i)_{i=1}^N$. Unless the data has time-series nature, it is customary to
presume independence of samples $z_i$ conditional on $\omega$.
%
In supervised settings, the nature of the structure of each observation is also conjectured:
$z_i = (x_i, y_i)$ is typically assumed in regression or classification tasks with $
  p(z_i \mid \omega) = p(y_i \mid x_i, \omega) p(x_i)
$, i.e. the inputs $x_i$ are produced by some opaque not particularly interesting process.
%
In unsupervised settings a hierarchical model is usually considered: it is hypothesized that
there are unobserved causal variables $u_i$ with some prior distribution $\pi(u_i \mid \omega)$
and a model of $p(x_i \mid u_i, \omega)$.
% vae x -- indep: q(u \mid x) ~ p(u \mid x, D), q(\omega \mid D) see \cite[eq.~4]{1805.08498}
%
In following we assume that the samples in $D$ are conditionally independent.

The posterior distribution $p(\omega \mid D)$ derived by Bayesian inference from the premise,
can be used to get useful information about yet unobserved data, such as the classification
uncertainty and predictive statistics, to name some examples. These downstream tasks usually
involve iterated or nested integrals over the posterior, e.g. classification uncertainty
evaluates the entropy $
  \mathbb{H}(q) = - \mathbb{E}_{y\sim q} \log{q(y)}
$ of the posterior predictive distribution at $
  p(y \mid x) = \mathbb{E}_{p(\omega \mid D)} p(y \mid x, \omega)
$ for some $x$ of interest, whereas the predictive statistics use
$
  \mathbb{E}_{p(\omega \mid D)} \hat{g}_\omega(x)
$ (by Fubini) for $
  \hat{g}_\omega(x) = \mathbb{E}_{p(y \mid x, \omega)} g(y)
$. These integrals are rarely analytically tractable in practice, and so are approximated
using Monte-Carlo estimators with varying degree of precision, \cite{citation_needed}.

However, despite the modeling assumption being chosen by the practitioner, the posterior
itself can be intractable or scale poorly with the problem size, due to the integral in the
denominator in the Bayes rule. Thus, save for the relatively simple cases \cite{citation_needed},
the exactness of Bayesian Inference is traded for tractability and speed of Variational
Inference.

The main idea of Variational Inference is to recast the posterior inference problem into
optimization framework. Instead of deriving $p(\omega \mid D)$ exactly, we consider a parametric
family $\mathcal{Q}$ of distributions indexed by a variational parameter $\theta \in \Theta$
as potential approximations for the posterior. Here $\theta$ represents a generic composite
parameter of the approximation and is specified whenever necessary.

We seek the closest $Q_\theta(d\omega)$ within the class with respect to some distance or
proximity score $\rho$ on probability measures or densities. Thus the class $\mathcal{Q}$ becomes
part of the modeling assumptions and is usually picked so that its members can be sampled
from, have known and tractable density $
  q_\theta(\omega) d\omega
$ with a tractable score function $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$ for REINFORCE, \cite{citation_needed}, or amenable to the Reparameterization Trick or
its variants, \cite[sec.~2.4]{kingma_auto-encoding_2014}, \cite[sec.~3]{figurnov_implicit_2019}: $
  \omega \sim q_\theta(\omega)
$
is equivalent in distribution to $
  \omega = g_\theta(\varepsilon)
$ for some deterministic differentiable $
  \theta \mapsto g_\theta(\varepsilon)
$ and a non-parametric random variable $
  \varepsilon \sim p_\varepsilon(\varepsilon)
$.
% (reparameterization trick) $q_{\theta}(d\omega)$ is a push-forward of $p(d\varepsilon)$ by a
% differentiable map $g_\theta(\varepsilon)$.

Not every proximity score is practically useful, since finding the variational parameters
$$
\theta^*
  \in \arg \min_{\theta} \rho\bigl(
    q_\theta(\omega), p(\omega \mid D)
  \bigr)
  \,, $$
still requires access to the exact posterior, precisely what variational inference attempts
to avoid. Despite its shortcomings, such as failing to be a metric, and infinite values for
densities with non-nested support, \cite{citation_needed}, the Kullback-Leibler divergence,
defined for $q(\omega) \ll p(\omega)$ as
$$
KL(q\| p)
  = \mathbb{E}_{\omega \sim q}
    \log\frac{q(\omega)}{p(\omega)}
  \,, $$
is commonly used. It circumvents the raised issue, because it satisfies the following identity,
which holds for $q$ and $p$ whenever the divergences are finite:
\begin{equation} \label{eq:kl-div-master}
  \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q(\omega) \| \pi(\omega))
    + KL(q(\omega) \| p(\omega \mid D))
    \,.
\end{equation}
This identity implies an equivalent proxy problem for the variational inference -- maximizing
the \textit{evidence lower bound} (ELBO)
\begin{equation} \label{eq:elbo_general}
\begin{aligned}
  & \underset{\theta}{\text{maximize}}
    & \mathcal{L}(\theta; \pi, p)
      & = \mathbb{E}_{\omega \sim q_{\theta}}
          \log p(D \mid \omega)
        - KL(q_{\theta}(\omega) \| \pi(\omega))
        % \mathbb{E}_{\omega \sim q_{\theta}}
          % \log\frac{q_\theta(\omega)}{\pi(\omega)}
  % \\
  % & &
  %     & = \mathbb{E}_{\omega \sim q_{\theta}}
  %         \log{p(D \mid \omega) \pi(\omega)}
  %       + \mathbb{H}(q_{\theta})
  \,,
\end{aligned}
\end{equation}
where the prior and conditional data probability model as listed as arguments, to emphasize
that they can potentially be optimized over (using alternating approach leading to EM-algorithm).
% vae
It is worth noting, that the optimization objective \eqref{eq:elbo_general} requires only $
  p(\omega \mid D)
    \propto p(D \mid \omega) \pi(\omega)
$, i.e. up to its normalizing constant, which is crucial for its practical viability. As
was mentioned above, other objectives for variational approximation can be considered, e.g.
\cite{ranganath_operator_2018} propose an approach that encompasses $f$-divergences, \textbf{...} and
other proximity scores, that depend on $p(\omega\mid D)$ through $p(D \mid \omega)$ and
$\pi(\omega)$.

The Stochastic Gradient Variational Bayes, proposed by \cite{kingma_auto-encoding_2014,citation_needed}, is
an unbiased differentiable Monte-Carlo estimator of the lower bound \eqref{eq:elbo_general}
that allows maximizing ELBO using stochastic gradient-based methods. In deep Bayesian inference
the variational approximation family is defined by deep neural networks, using the stochastic
estimate of ELBO the only viable option. Exploiting independence of observations in $D$ and
the reparameterization trick, the estimator is defined as
\begin{equation} \label{eq:elbo}
  \mathcal{L}^{SGVB}(\theta; \pi, p)
    =
    - KL(q_{\theta}(\omega) \| \pi(\omega))
    % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)
    + N
      \frac1{M} \sum_{i=1}^M
        \frac1{L} \sum_{l=1}^L
          \log p(z_i \mid \omega_{il})
            \Big\vert_{\omega_{il} = g_\theta(\varepsilon_{il})}
    % =
    % - KL(q_{\theta}(\omega) \| \pi(\omega))
    % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
    %   \hat{\mathbb{E}}_{z,\omega \sim B}
    %     \log p(z \mid \omega)
    \,, 
\end{equation}
where $B = (z_i)_{i=1}^M$ is a minibatch form the dataset $D$ and $
  (\varepsilon_{il})_{l=1}^L \sim p(\varepsilon)
$ are drawn independently per each sample $z_i$. The most commonly used variational
approximations have analytically tractable KL-divergence term, but can be replaced by
an unbiased differentiable estimator.

For a particular family of variational posterior approximation, variational inference
can be used as a regularization method \cite{kingma_variational_2015} and as model sparsification
technique \cite{molchanov_variational_2017}. In \cite{kingma_variational_2015} the Binary
dropout and DropConnect \cite{hinton_improving_2012,dropconnect}, and Gaussian dropout
\cite{srivastava_dropout_2014,wang_fast_2013} were reinterpreted as variational inference
methods. In that study, authors argue that dropout, being multiplicative parameter noise,
induces parameter distribution equivalent to a fully factorized variational posterior
approximation of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where each component $q_{\theta}(\omega_j)$ is defined via $\omega_j = \mu_j \xi_j$
with $\xi_j \sim p(\xi_j)$ iid for some $p(\xi)$. In Binary dropout with rate $p \in (0, 1)$
the noise $\xi$ is binary and distributed $
  \mathcal{B}\bigl(
      \{0, \tfrac1{1-p}\}, 1-p
    \bigr)
$, whereas in Gaussian dropout $
  \xi \sim \mathcal{N}(1, \alpha)
$ for $\alpha = \tfrac{p}{1-p}$. In the latter case, the \textit{dropout score} $\alpha$ acts
like a variance scaling factor, making it a natural variational parameter for the posterior
approximation, which is optimized via SGVB \eqref{eq:elbo}. Thus \cite{kingma_variational_2015}
propose Variational Dropout technique for Bayesian model regularization. The flexibility in
specifying dependence of $q_\theta$ on $\alpha$ allows imposing structure on dropout: global
rate for the whole model network, local per each component and even specific to individual
parameter.

Secondary contribution of \cite{kingma_variational_2015} has been the analysis of the local
reparameterization trick as a gradient variance reduction technique for \eqref{eq:elbo},
that is particularly suitable for Variational Dropout. The trick is to translate uncertainty
from the global parameter noise to local noise in intermediate outputs within the network.
Initially, this idea was used in \cite{who-dis-originally-22} for Gaussian dropout,
who employed the closure of Gaussians under affine transformations as a way to inject
regularizing noise. Later in \cite[app.~B]{kingma_variational_2015} it was rigorously shown
that for the case of fully factorized Gaussian dropout posterior approximations, the
gradient estimator $\nabla_\theta \mathcal{L}^{SGVB}$ is more statistically efficient
and computationally fast, when the trick is used in addition to independent per sample
parameter draws, which reduce variance by decorrelating ELBO estimates within the minibatch.

@@

Under the original $(\mu, \alpha)$ parameterization the stochastic output of a linear
layer $
  y = W^\top x + b
$ with $
  W \in \mathbb{R}^{n\times m}
$ and independent $
  W_{ij} \sim \mathcal{N}(\mu_{ij}, \alpha_{ij} \mu_{ij}^2)
$ is equivalent in distribution to $
  y \sim \prod_i \mathcal{N}\Bigl(
      \sum_j \mu_{ij} x_j + b_i,
      \sum_j \alpha_{ij} \mu_{ij}^2 x_j^2
  \Bigr)
$. \cite{molchanov_variational_2017} argue that for the purposes of model sparsification
this parameterization should be abandoned in favour of an equivalent $(\mu, \sigma^2)$ with
the score $\alpha$ calculated using $\tfrac{\sigma^2}{\mu^2}$ when needed. This makes $
  \nabla_\mu \mathcal{L}^{SGVB}
$, which is particularly important for sparsification, independent of the noise injected
by the trick.

@@

@@

Let $(\mu_i)_{i=1}^m \in \cplx$ and $
  (z_i)_{i=1}^m \sim \mathcal{N}^{\cplx}(0, 1, 0)
$ iid. Then $W = \sum_i \lvert \theta_i + z_i \rvert^2$ has a non-central $\chi^2_{2m}$
distribution with non-centrality parameter $s^2 = \sum_i \lvert \theta_i \rvert^2$. The
log-moments of $W$ for general integer $m \geq1$ are given in
\cite[p.~2466]{lapidoth_capacity_2003}.  % in fact Appendix X, Lemma 10.1
In particular, for $m=1$ and $\theta\in \cplx$ we have
\begin{equation}  \label{eq:log-moment-for-chi-2}
\mathbb{E}_{z \sim \mathcal{N}^{\cplx}(0, 1, 0)}
  \log \lvert \theta + z \rvert^2
  = \log \lvert \theta \rvert^2 - \mathop{Ei}( - \lvert \theta \rvert^2)
  \,,
\end{equation}
where $\mathop{Ei}(x) = \int^x_{-\infty} t^{-1} e^t dt$ for $x < 0$ is the exponential integral
function. Note that the limiting value of \eqref{eq:log-moment-for-chi-2} as $
  \lvert \theta \rvert \to 0
$ is $- \gamma$, i.e. negative Euler's constant. Therefore after cancellation the expression
of the KL-divergence in \eqref{eq:var-do-kl-div-raw} becomes
\begin{equation}  \label{eq:var-do-kl-div}
KL(q\| \pi)
  \propto
  %   - \log \lvert \det{(\pi e \sigma^2)} \rvert
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 0, 1, 0)}
  %       \log{\bigl\lvert \tfrac1{\sqrt{\alpha}} + \xi \bigr\rvert^2}
  % =
  %   - \log{\sigma^2}  % - \log{\pi e}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \bigl(
  %     - \log \alpha - \mathop{Ei}( -\tfrac1{\alpha})
  %   \bigr)
  % =
    % \log \tfrac{\lvert \mu \rvert^\beta}{\sigma^2}
    % \log \tfrac{\lvert \mu \rvert^\beta}{\alpha \lvert \mu \rvert^2}
    \tfrac{\beta-2}2 \log{\lvert \mu \rvert^2}
    + \log{\tfrac1{\alpha}}
    - \tfrac{\beta}2 \mathop{Ei}(- \tfrac1{\alpha})
  \,.
\end{equation}

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.47\textwidth}
    \includegraphics[draft]{../notebooks/assets/curve_fit.pdf}
  \end{subfigure}%\hspace{-2.em}
  \begin{subfigure}[b]{0.47\textwidth}
    \includegraphics[draft]{../notebooks/assets/grad_log.pdf}
  \end{subfigure}
  \caption{Fit of the regression \cite{molchanov_variational_2017}, and the accuracy of
  the derivative wrt $\log{\alpha}$.}
  \label{fig:molchanov-derivative-replica}
\end{figure}

>>>>

Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge. Solving the storage and compute constraints encourages development
of model compress and sparsification methods, that focus on favourable tradeoff between
performance and size. Most popular techniquees include weight pruning \cite{citation_needed},
model quantization from IEEE754 floating point to integer-based arithmetic \cite{citation_needed},
and dropout. The key assumption is that the more explicitly zero parameters, the less
computations is needed.

Bayesian Inference is a principled framework of reasoning about uncertainty. The core principle
is updating prior beliefs in accordance with the likelihood of empirical observations into
posterior belief. Under suitable modelling and prior assumptions, Bayesian methods can be
used towards model sparsification, \cite{kingma_variational_2015,molchanov_variational_2017},
and \cite{kharitonov_variational_2018}.

@@

In general, the core of Bayesian Inference can be summarized as follows: given a prior
distribution $\pi(m)$ on models (hypotheses) $\mathcal{M}$, utilize the empirical evidence
$D$ to update the assumptions by considering the likelihood of the observations under each
$m \in \mathcal{M}$. Models are represented by a parametric family indexed by $\omega \in
\Omega$ and each $m_\omega(\cdot)$ specifies the conditional distribution of the data $
  D = (z_i)_{i=1}^N
$. The posterior distribution $p(\omega \mid D)$, derived using the Bayes rule $
  p(m \mid D) = \tfrac{p(D \mid m) \pi(m)}{p(D)}
$ with $
  p(D) = \mathbb{E}_{\pi(m)} p(D \mid m)
$, provides useful information about yet unobserved data and model parameters, e.g
classification uncertainty, predictive statistics, and parameter relevance.

The posterior itself is computationally intractable, save for the relatively simple cases
\cite{citation_needed}, and therefore exact inference is traded for tractability and scalability
of Variational Inference. Instead of deriving $p(\omega \mid D)$ using the rule, this approach
recasts posterior inference into variational optimization problem
\begin{equation} \label{eq:variational-progam}
  q_*
    \in \arg \min_{\theta} \rho\bigl(
      q_\theta(\omega), p(\omega \mid D)
    \bigr)
    \,,
\end{equation}
over a tractable parametric family of distributions on the parameter space $
  \mathcal{Q} = \{Q_\theta(d\omega) \colon \theta \in \Theta\}
$. The class $\mathcal{Q}$ is picked so that its member have tractable densities $
  q_\theta(\omega) d\omega
$, can be sampled from and possess tractable $\log$-derivatives $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$, \cite{williams_simple_1992}, or can be represented as differentiable transformations
of non-parametric noise, \cite{kingma_auto-encoding_2014,figurnov_implicit_2019}.
% (reparameterization trick, pathwise gradieent) $q_{\theta}(d\omega)$ is a push-forward
% of $p(d\varepsilon)$ by a differentiable map $g(\varepsilon; \theta)$.
% % http://stillbreeze.github.io/REINFORCE-vs-Reparameterization-trick/#fn:1

Kullback-Leibler divergence, defined for $q(\omega) \ll p(\omega)$ as
$$
KL(q\| p)
  = \mathbb{E}_{\omega \sim q}
    \log\frac{q(\omega)}{p(\omega)}
  \,, $$
is the natural choice for $\rho$, despite it not being a metric and requiring distributions
with nested support. The main reason is that it satisfies the following identity, which when
everything works, holds for any $q$ and $p$:
\begin{equation} \label{eq:kl-div-master}
  \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q(\omega) \| \pi(\omega))
    + KL(q(\omega) \| p(\omega \mid D))
    \,.
\end{equation}
This key identity yield an equivalent problem for \eqref{eq:variational-progam}:
maximizing the \textit{evidence lower bound} (ELBO)
\begin{equation} \label{eq:elbo_general}
\begin{aligned}
  & \underset{\theta}{\text{maximize}}
    & \mathcal{L}(\theta; \phi, \lambda)
      & = \mathbb{E}_{\omega \sim q_{\theta}}
          \log p_{\phi}(D \mid \omega)
        - KL(q_{\theta}(\omega) \| \pi_{\lambda}(\omega))
        % \mathbb{E}_{\omega \sim q_{\theta}}
          % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % \\
  % & &
  %     & = \mathbb{E}_{\omega \sim q_{\theta}}
  %         \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %       + \mathbb{H}(q_{\theta})
  \,,
\end{aligned}
\end{equation}
where the prior and likelihood are allowed to depend on explicit parameters. The most
commonly used algorithm to optimize \eqref{eq:elbo_general} is to use EM-algorithm.
Other variational objectives are possible, provided $p(\omega \mid D)$ is probed only
through $\log p(D \mid \omega)$ and $\log \pi(\omega)$, \cite{ranganath_operator_2018}.

Stochastic Gradient Variational Bayes (SGVB), proposed by \cite{kingma_auto-encoding_2014},
is an differentiable unbiased Monte-Carlo estimator of \eqref{eq:elbo_general}, which makes
it possible to employ stochastic gradient-based optimization methods. If $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$ and differentiable $
  g(\varepsilon; \theta)
$, then the SGVB is
\begin{equation} \label{eq:sgvb}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta}(\omega) \| \pi_{\lambda}(\omega))
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{l=1}^L
        \log p_{\phi}(D \mid \omega_{il})
          \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta}(\omega) \| \pi(\omega))
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,, 
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$. Intractable KL-divergence terms can be also
replaced by differentiable unbiased Monte-Carlo estimators, \cite{kingma_auto-encoding_2014}.
%
For the purposes of this study, we assume that the observed data is independent, and
use minibatch version of \eqref{eq:sgvb} with $L=1$:
\begin{equation} \label{eq:elbo}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta}(\omega) \| \pi_{\lambda}(\omega))
      + \frac{N}{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,,
\end{equation}
for a random subsample $(z_i)_{i=1}^M$ from $D$ and iid $
  (\varepsilon_i)_{i=1}^M \sim p_\varepsilon
$.

\subsection{Dropout} % (fold)
\label{sub:dropout}

For a particular family of posterior approximation, variational inference can be used
as a regularization method \cite{kingma_variational_2015} and as model sparsification
technique \cite{molchanov_variational_2017}. In \cite{kingma_variational_2015} Binary
dropout, DropConnect and Gaussian dropout, \cite{hinton_improving_2012,wan_regularization_2013},
\cite{srivastava_dropout_2014}, and \cite{wang_fast_2013} were reinterpreted as variational
inference methods. That study argues that the multiplicative noise, introduced by dropout,
induces a distribution equivalent to a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $\xi_j \sim p(\xi_j)$
iid from some $p_\xi$. In Binary dropout with rate $p \in (0, 1)$ we have $
  p_\xi
    = \mathcal{B}\bigl(
        \{0, \tfrac1{1-p}\}, 1-p
      \bigr)
$, whereas in Gaussian dropout $
  \xi \sim \mathcal{N}(1, \alpha)
$ for $\alpha = \tfrac{p}{1-p}$. In the latter case, the \textit{dropout score} $\alpha$ acts
like a variance scaling factor, making it a natural variational parameter for the posterior
approximation, which is optimized via SGVB \eqref{eq:elbo}. Thus \cite{kingma_variational_2015}
propose Variational Dropout technique for Bayesian model regularization. The flexibility in
specifying dependence of $q_\theta$ on $\alpha$ allows imposing structure on dropout: global
rate for the whole model network, local per each component and even specific to individual
parameter.

Secondary contribution of \cite{kingma_variational_2015} has been the analysis of the local
reparameterization trick as a gradient variance reduction technique for \eqref{eq:elbo},
that is particularly suitable for Variational Dropout. The trick is to translate uncertainty
from the global parameter noise to local noise in intermediate outputs within the network.
Initially, this idea was used in \cite{who-dis-originally-22} for Gaussian dropout,
who employed the closure of Gaussians under affine transformations as a way to inject
regularizing noise. Later in \cite[app.~B]{kingma_variational_2015} it was rigorously shown
that for the case of fully factorized Gaussian dropout posterior approximations, the
gradient estimator $\nabla_\theta \tilde{\mathcal{L}}$ is more statistically efficient
and computationally fast, when the trick is used in addition to independent per sample
parameter draws, which reduce variance by decorrelating ELBO estimates within the minibatch.

@@

Under the original $(\mu, \alpha)$ parameterization the stochastic output of a linear
layer $
  y = W^\top x + b
$ with $
  W \in \mathbb{R}^{n\times m}
$ and independent $
  W_{ij} \sim \mathcal{N}(\mu_{ij}, \alpha_{ij} \mu_{ij}^2)
$ is equivalent in distribution to $
  y \sim \prod_i \mathcal{N}\Bigl(
      \sum_j \mu_{ij} x_j + b_i,
      \sum_j \alpha_{ij} \mu_{ij}^2 x_j^2
  \Bigr)
$. \cite{molchanov_variational_2017} argue that for the purposes of model sparsification
this parameterization should be abandoned in favour of an equivalent $(\mu, \sigma^2)$ with
the score $\alpha$ calculated using $\tfrac{\sigma^2}{\mu^2}$ when needed. This makes $
  \nabla_\mu \tilde{\mathcal{L}}
$, which is particularly important for sparsification, independent of the noise injected
by the trick.

@@

% subsection dropout (end)

@@

Let $(\mu_i)_{i=1}^m \in \cplx$ and $
  (z_i)_{i=1}^m \sim \mathcal{N}^{\cplx}(0, 1, 0)
$ iid. Then $W = \sum_i \lvert \theta_i + z_i \rvert^2$ has a non-central $\chi^2_{2m}$
distribution with non-centrality parameter $s^2 = \sum_i \lvert \theta_i \rvert^2$. The
log-moments of $W$ for general integer $m \geq1$ are given in
\cite[p.~2466]{lapidoth_capacity_2003}.  % in fact Appendix X, Lemma 10.1
In particular, for $m=1$ and $\theta\in \cplx$ we have
\begin{equation}  \label{eq:log-moment-for-chi-2}
\mathbb{E}_{z \sim \mathcal{N}^{\cplx}(0, 1, 0)}
  \log \lvert \theta + z \rvert^2
  = \log \lvert \theta \rvert^2 - \mathop{Ei}( - \lvert \theta \rvert^2)
  \,,
\end{equation}
where $\mathop{Ei}(x) = \int^x_{-\infty} t^{-1} e^t dt$ for $x < 0$ is the exponential integral
function. Note that the limiting value of \eqref{eq:log-moment-for-chi-2} as $
  \lvert \theta \rvert \to 0
$ is $- \gamma$, i.e. negative Euler's constant. Therefore after cancellation the expression
of the KL-divergence in \eqref{eq:var-do-kl-div-raw} becomes
\begin{equation}  \label{eq:var-do-kl-div}
KL(q\| \pi)
  \propto
  %   - \log \lvert \det{(\pi e \sigma^2)} \rvert
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 0, 1, 0)}
  %       \log{\bigl\lvert \tfrac1{\sqrt{\alpha}} + \xi \bigr\rvert^2}
  % =
  %   - \log{\sigma^2}  % - \log{\pi e}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \bigl(
  %     - \log \alpha - \mathop{Ei}( -\tfrac1{\alpha})
  %   \bigr)
  % =
  %   - \log{\sigma^2}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   - \tfrac{\beta}2 \log \alpha
  %   - \tfrac{\beta}2 \mathop{Ei}( -\tfrac1{\alpha})
  % =
    % \log \tfrac{\lvert \mu \rvert^\beta}{\sigma^2}
    % \log \tfrac{\lvert \mu \rvert^\beta}{\alpha \lvert \mu \rvert^2}
    \tfrac{\beta-2}2 \log{\lvert \mu \rvert^2}
    + \log{\tfrac1{\alpha}}
    - \tfrac{\beta}2 \mathop{Ei}(- \tfrac1{\alpha})
  \,.
\end{equation}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=1.\linewidth]{../notebooks/assets/grad_log.pdf}
  \caption{$\tfrac{d f(\alpha)}{d \log{\alpha}}$ of the fit \cite{molchanov_variational_2017},
  MC estimate, and exact derivative using \eqref{eq:real-kl-div-deriv-log}.}
  \label{fig:molchanov-derivative-replica}
\end{figure}
